\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-09 k-planar.tex

  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to $V(F)$.
   %
   %
   % If $k=3$, set $R_i := P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citep{DJMMUW20}, split $P_1,\ldots,P_k$ to partition $V(F)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a non-empty path and each $R_i$ contains vertices from exactly one of $P_1,\ldots,P_k$.
  %
	% Next, as in \citep{DJMMUW20}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $N'-V(F)$.
	Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $\overline{Y}^+$ denote the subgraph of $N$ consisting of the vertices and edges of $\overline{Y}$ plus the vertices and edges of every kite formed by a crossing between an edge of $G$ and an edge of $\overline{Y}$.  Let $Y:= V(\overline{Y}-F)$ and let $Y^+:=V(\overline{Y}^+-F)$.

	We claim that, for each integer $i\ge 0$, $|V(\overline{Y}^+)\cap L_i|\le 15$.  First observe that, since $Q_1,Q_2,Q_3$ are each vertical paths in $T$,  $\overline{Y}$ contains at most three vertices of $L_i$, each incident on at most two edges of $\overline{Y}$.  Since $\dist_{G'}(v,w)\le 2$ for each $vw\in E(G)$, any vertex $x\in V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i$, is incident to an edge $xy\in E(G)$ that crosses one of the at most six edges in $\overline{Y}$ having an endpoint in $L_i$.  These at most six edges have at most 12 endpoints.  Therefore $|V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i|\le 6\times 2=12$, so $|V(\overline{Y}^+)\cap L_i|\le 12+3=15$.

	Let $M$ and $M^+$ denote the subgraph of $G$ containing the edges and vertices of $\overline{Y}$, respectively $\overline{Y}^+$, and the edges and vertices of $F$. (See \cref{boring_figure}.)  By our choice of $\tau$, the graph $M$ is $2$-connected.  Furthermore, the vertices on the boundary of each inner face of $M$ are contained in $Y\cup P_a\cup P_b$ for some $a,b\in\{1,\ldots,k\}$.  In particular if $k=3$ then, for each inner face $F_i$ of $M$ there exists some $c\in\{1,2,3\}$ such that no vertex of $P_c$ is on the boundary of $F_i$.

  \begin{figure}
    \begin{center}
      \begin{tabular}{c@{\hspace{1cm}}c}
        \includegraphics{figs/mmplus-1} & \includegraphics{figs/mmplus-2} \\
        $M$ & $M^+$
      \end{tabular}
    \end{center}
    \caption{The graphs $M$ and $M^+$.}
    \label{boring_figure}
  \end{figure}

  The graph $M^+$ has some number of inner faces, all contained in the interior of $F$. Some of the inner faces of $M^+$ are Type~(ii) faces of $G$. Let $F_1,\ldots,F_m$ be the inner faces of $M^+$ that are not Type~(ii) faces of $G$.  We claim that, for each $i\in\{1,\ldots,m\}$, the boundary of $F_i$ is a cycle in $G'$ that contains no spars. Otherwise, some edge $vw$ contributes to the boundary of $F_i$ but is crossed by an edge $xy\in E(G)$. Then, $vw\not\in E(F)$ since no edge of $F$ is crossed by any edge of $G$. Therefore $vw\in E(\overline{Y}^+)$ so $xy\in E(\overline{Y}^+)$. But then the only faces of $M^+$ with $vw$ on their boundary are Type~(ii) faces.  In particular $vw$ cannot be on the boundary of $F_i$.

  Observe that each of the faces $F_1,\ldots,F_m$ is contained in a single inner face of $M$.  Therefore, $V(F_i)$ can be partitioned into at most three sets $P_1'$, $P_2'$, and $P_3'$ where $P_1'\subset V(Y^+)$, $P_2'\subseteq P_a$, $P_3'\subseteq P_b$ for some $a,b\in\{1,2,3\}$, and $F_i[P_j']$ is a path, for each $j\in\{1,2,3\}$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_1',P_2',P_3'$ of $V(F_i)$ to obtain the desired $H$-partition $\PP_i$ and tree decomposition of subgraph $N_i$ of $G$ contained in $F_i$.

  \note{PM}{Pick up here tomorrow afternoon. For now I have to teach and prepare a talk.}
  The proof finishes in the same way as the proof in \cite{DJMMUW20}.  The paths $P_1,\ldots,P_k$, and $S=V(Y^+)$ become elements of the $H$-partition.
  Elements in each of the $H$-partitions of $N_1,\ldots,N_m$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.

>>>>

  For each $i\in\{1,2,3\}$, let $Q_i$ be the shortest path, in $T$, from $v_i$ to $V(F)$.
   %
   %
   % If $k=3$, set $R_i := P_i$ for each $i\in\{1,2,3\}$.  Otherwise, as in \citep{DJMMUW20}, split $P_1,\ldots,P_k$ to partition $V(F)$ into three sets $R_1$, $R_2$, and $R_3$ such that each $F[R_i]$ is a non-empty path and each $R_i$ contains vertices from exactly one of $P_1,\ldots,P_k$.
  %
	% Next, as in \citep{DJMMUW20}, use Sperner's Lemma to find an inner face $\tau=v_1v_2v_3$ of $N'$ such that, $T$ contains disjoint vertical paths $Q_1,Q_2,Q_3$ such that each $Q_i$ begins at $v_i$, ends at some vertex in $R_i$, and whose internal vertices (if any) are contained in $N'-V(F)$.
  Let $\overline{Y}$ denote the subgraph of $N'$ consisting of vertices and edges of $Q_1$, $Q_2$, $Q_3$, and $\tau$.  Let $S:=V(Q_1)\cup V(Q_2)\cup V(Q_3)$ and let $v:=v_p$.  Observe that, for each $i\in\{1,2,3\}$ and each $j\ge 0$, $|V(Q_i)\cap L_j|\le 1$.  Therefore $|S\cap L_j|\le 3$ for each integer $j\ge 0$.  Therefore, taking $Y:=S\cup\{v\}\setminus V(F)$ as a part in the partition $\PP$ will satisfy Requirement~2 of the lemma.

	% We claim that, for each integer $i\ge 0$, $|V(\overline{Y}^+)\cap L_i|\le 15$.  First observe that, since $Q_1,Q_2,Q_3$ are each vertical paths in $T$,  $\overline{Y}$ contains at most three vertices of $L_i$, each incident on at most two edges of $\overline{Y}$.  Since $\dist_{G'}(v,w)\le 2$ for each $vw\in E(G)$, any vertex $x\in V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i$, is incident to an edge $xy\in E(G)$ that crosses one of the at most six edges in $\overline{Y}$ having an endpoint in $L_i$.  These at most six edges have at most 12 endpoints.  Therefore $|V(\overline{Y}^+)\setminus V(\overline{Y})\cap L_i|\le 6\times 2=12$, so $|V(\overline{Y}^+)\cap L_i|\le 12+3=15$.

	Let $M$ denote the subgraph of $G'$ containing the edges and vertices of $\overline{Y}$ and the edges and vertices of $F$. (See \cref{boring_figure}.)  By our choice of $\tau$, the graph $M$ is $2$-connected.  Furthermore, the vertices on the boundary of each inner face of $M$ are contained in $Y\cup P_a\cup P_b$ for some $a,b\in\{1,\ldots,k\}$.  In particular if $k=3$ then, for each inner face $F_i$ of $M$ there exists some $c\in\{1,2,3\}$ such that no vertex of $P_c$ is on the boundary of $F_i$.

  \begin{figure}
    \begin{center}
      \begin{tabular}{c@{\hspace{1cm}}c}
        \includegraphics{figs/mmplus-1} & \includegraphics{figs/mmplus-2} \\
        $M$ & $M^+$
      \end{tabular}
    \end{center}
    \caption{The graphs $M$ and $M^+$. \note{PM}{This figure needs updating now.}}
    \label{boring_figure}
  \end{figure}

  The graph $M$ has some number of inner faces $F_1,\ldots,F_m$, each of which is contained in the interior of $F$. 
  % Some of the inner faces of $M^+$ are Type~(ii) faces of $G$. Let $F_1,\ldots,F_m$ be the inner faces of $M^+$ that are not Type~(ii) faces of $G$.  We claim that, for each $i\in\{1,\ldots,m\}$, the boundary of $F_i$ is a cycle in $G'$ that contains no spars. Otherwise, some edge $vw$ contributes to the boundary of $F_i$ but is crossed by an edge $xy\in E(G)$. Then, $vw\not\in E(F)$ since no edge of $F$ is crossed by any edge of $G$. Therefore $vw\in E(\overline{Y}^+)$ so $xy\in E(\overline{Y}^+)$. But then the only faces of $M^+$ with $vw$ on their boundary are Type~(ii) faces.  In particular $vw$ cannot be on the boundary of $F_i$.
  Each of the faces $F_1,\ldots,F_m$ is contained in a single inner face of $M$.  Therefore, $V(F_i)$ can be partitioned into at most three sets $P_1'$, $P_2'$, and $P_3'$ where $P_1'\subset V(Y)$, $P_2'\subseteq P_a$, $P_3'\subseteq P_b$ for some $a,b\in\{1,2,3\}$, and $F_i[P_j']$ is a path, for each $j\in\{1,2,3\}$.  Therefore, we can apply induction using the cycle $F_i$ and the partition $P_1',P_2',P_3'$ of $V(F_i)$ to obtain the desired $H$-partition $\PP_i$ and tree decomposition of subgraph $N_i$ of $G$ contained in $F_i$.

  \note{PM}{Pick up here tomorrow, after teaching.}
  The proof finishes in the same way as the proof in \cite{DJMMUW20}.  The paths $P_1,\ldots,P_k$, and $S=V(Y^+)$ become elements of the $H$-partition.
  Elements in each of the $H$-partitions of $N_1,\ldots,N_m$ that intersect $P_1,\ldots,P_k$, or $V(\overline{Y}^+-F)$ are discarded and all the resulting sets are combined to obtain an $H$-partition of $G$.
